---
title: "Prosper Loan Analysis"
author: "Luca Miniati"
output: html_document
---

## Background
Prosper is a peer-to-peer lending platform where borrowers can find investors to fund their loans.
Prosper's business relies on the idea of fractionalization, where funds for loans are
essentially "crowd sourced" from investors, who can invest as little as $25 per loan.

I will be analyzing the performance of Prosper loans from 2023-2024. The dataset is availble at
<https://help.prosper.com/hc/en-us/articles/210013083-Where-can-I-download-data-about-loans-through-Prosper>
with a free account.

For the subsequent analysis, it's important to define some terms.  

- Peer-to-peer lending: A system that connects borrowers directly to investors, cutting out banks
  from the process. 
- Default: If a borrower fails to pay back a loan, the loan is considered "defaulted". The
  borrower's credit score is impacted, and the investors are not compensated for their loss.  
- Completion: If a borrower pays the loan back in full, the loan is considered "completed".  
- Charge-off: If a borrower fails to pay multiple payments on a loan, the financial
  institution that created the loan is likely to write the loan off as a loss, essentially
  expecting it to default. Borrower still technicaly owes the money, but the debt might be sold to
  a new buyer or transferred to a collection agency.  
- Charge-off recovery: In the event of a charge-off, the borrower still has a chance to repay the
  loan. If they do, the loan is considered "recovered".
- Status: The current status of a loan (e.g "Defaulted", "Completed", etc.).  
- Amount: The amount of money the borrower takes from the investors.  
- Term: How long the borrower has to repay the amount.  
- Annual Interest Rate (APR): The interest rate that is charged to the borrower each year. All
  Prosper loans are [amortized](https://www.investopedia.com/terms/a/amortized_loan.asp).  

```{r include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(fig.width=10)
options(width = 100)
library(tidyverse)
library(lubridate)
```

## Tidying
```{r}
df <- read.csv("data/data.csv")

nrow(df[duplicated(df), ])
```

At a first glance, Prosper processed 135809 unique loans in 2023, tracking 22 columns of data with
no duplicate rows.

```{r}
# Check for null values
colSums(is.na(df))
```

In the dataset, it looks like there aren't any null values except in the `loan_default_reason`
column, which has an unusally high number of null values (135621 values out of the 135809 total
rows). This is because only a fraction of the loans are actually defaulted; the rest are current
(in progress) or completed (without default). These other loans don't have a default reason in the
table because they aren't defaulted.

```{r}
# Selecting the loans that have a non-null default reason
loans_with_default_reason <- df[!is.na(df$loan_default_reason), ]
head(loans_with_default_reason)
```

In the results of the loans with a default reason, there are some rows with
`loan_default_reason_description=DECEASED`. If someone passed away with an active loan, their loan
was marked as defaulted. From a financial standpoint, this makes sense, because a death would be
considered a financial loss by the investor who gave them money
Because of this, we won't impute or drop these null value; rather, we'll treat the missing values
as meaningful in indicating non-defaulted loans.

```{r}
# Check data types of each column
sapply(df, class)
```

The data types look mostly correct, with the exception of `origination_date` and `prosper_rating`.  

```{r}
df$origination_date <- as.POSIXct(df$origination_date)
df$prosper_rating <- as.factor(df$prosper_rating)
```

For easier manipulation of dates, `origination_date` was converted into a POSIXct object.
Additionally, `prosper_rating` was converted to a factor to simplify the categorical analysis later.

```{r}
# Summary statistics for numeric variables
summary(df[, sapply(df, is.numeric)])
```

For the numeric variables, everything looks reasonable except for `loan_number` and
`loan_default_reason`. `loan_number` is a unique id given to each loan by Prosper. For the purposes
of this analysis, we'll drop the column, as it doesn't contribute anything. `loan_default_reason` is
a column of categorical features corresponding to the reason for a loan defaulting. For example,
a `loan_default_reason` of `1` corresponds to a borrower committing an unacceptable number of
delinquencies (late or missing payments). This directly corresponds to the human-readable label
column of `loan_default_reason_description`, so we'll drop `loan_default_reason` in favor of
`loan_default_reason_description`.

```{r}
df <- subset(df, select = -c(loan_number, loan_default_reason_description))
```

With our data tidy, we can move on.

## Exploratory Data Analysis

First, let's have an overview of the numeric features of the dataset.

```{r}
pairs(df[, sapply(df, is.numeric)])
```

From a birds-eye view, there seems to be slight relationships between some variables, but the data
is mostly scattered.

Next, I'd like to explore the aforementioned outliers.

```{r}
# Histograms
par(mfrow=c(1,3))
hist(df$days_past_due, main="Days Past Due", xlab="Days")
hist(df$prosper_fees_paid, main="Prosper Fees Paid", xlab="Fees")
hist(df$late_fees_paid, main="Late Fees Paid", xlab="Fees")
```

Most people pay back their loans on time, with some extreme outliers.

Prosper rates each individual borrower based on their credit, income, and past delinquencies. This
metric is used to grade borrowers on trustworthiness. This `prosper_rating` is a crucial field for
determining which loans are funded by investors, as it is blatantly displayed on the listings for
each loan on the Prosper website.

`late_fees_paid`.

```{r}
# Map `prosper_rating` to numberic values
ratings_numeric <- c("HR" = 1, "E" = 2, "D" = 3, "C" = 4, "B" = 5, "A" = 6, "AA" = 7)

# Convert prosper_rating to numeric using the mapping
df$prosper_rating_numeric <- ratings_numeric[as.character(df$prosper_rating)]

# Scatter plots
par(mfrow=c(1, 3))
plot(df$prosper_rating_numeric, df$days_past_due, main="Days Past Due vs. Prosper Rating", xlab="Prosper Rating", ylab="Days Past Due")
plot(df$prosper_rating_numeric, df$prosper_fees_paid, main="Prosper Fees Paid vs. Prosper Rating", xlab="Prosper Rating", ylab="Prosper Fees Paid")
plot(df$prosper_rating_numeric, df$late_fees_paid, main="Late Fees Paid vs. Prosper Rating", xlab="Prosper Rating", ylab="Late Fees Paid")


# Check the correlation
cor(df$prosper_rating_numeric, df$days_past_due)
cor(df$prosper_rating_numeric, df$prosper_fees_paid)
cor(df$prosper_rating_numeric, df$late_fees_paid)
```

It looks like their rating wasn't able to capture the volatile nature of peer-to-peer loan
payments. Let's see if it was able to predict loan status.

```{r}
# Filter the dataframe to include only final loan statuses (non-current loans)
df_filtered <- df[df$loan_status_description %in% c("DEFAULTED", "COMPLETED", "CHARGEOFF"), ]

# Calculate proportions of loan status by `prosper_rating`
loan_status_proportions <- prop.table(table(df_filtered$prosper_rating, df_filtered$loan_status_description),
                                      margin = 1)
# Convert to dataframe for plotting
loan_status_df <- as.data.frame(loan_status_proportions)

# Set data in correct order
loan_status_df <- loan_status_df %>%
    mutate(Var1 = factor(Var1, levels = c("AA", "A", "B", "C", "D", "E", "HR")),
           Var2 = factor(Var2, levels = c("DEFAULTED", "CHARGEOFF", "COMPLETED")))

# Plot stacked bar chart
ggplot(loan_status_df, aes(x = Var1, y = Freq, fill = Var2)) +
  geom_bar(stat = "identity") +
  scale_fill_manual(values = c("DEFAULTED" = "#F25F5C", "CHARGEOFF" = "#F0803C", "COMPLETED" = "#7DDE92")) +
  labs(x = "Prosper Rating", y = "Proportion (%)", fill = "Loan Status") +
  ggtitle("Loan Status Proportions by Prosper Rating") +
  theme_minimal()
```

While `prosper_rating` doesn't predict late payments well, it does seem to be a good predictor for
the final status of a loan. This makes sense because investors are likely to care less about when
they get paid, and more about whether they get their money back at all in the long run.

Peer-to-peer lending gets a bad rap because of its historical instability and high default rates.
Let's investigate Prosper's default rates in this dataset further.

```{r}
# Sort by `DEFAULTED` in descending order
loan_status_proportions <- loan_status_proportions[order(loan_status_proportions[, "DEFAULTED"]), ]
round(loan_status_proportions, digits = 4)
```

As expected, higher `prosper_rating` generally corresponds to a lower default rate. However, with
the criteria of default rate, `C` loans actually outperformed `B` loans and `E` loans outperformed
`D` loans. Let's investigate this further, since this isn't what we expected.

Taking a closer look, the discrepancy lies in the `CHARGEOFF` column. `C` loans have ~10% higher
rates of charge-off than those of `B` loans, which causes the shockingly low default rate.
Remember, loans that are charged off are likely to default. Prosper probably separates the rates of
charge-off and default in order to artificially deflate the default rate. Compared to the market
average default rate of greater than 10%
[[1]](https://jfin-swufe.springeropen.com/articles/10.1186/s40854-020-00197-y), all of Prosper's
default rates in 2023 were techically less than 1.7%, without factoring in charge-offs. The true
default rate is much higher if we consider charge-offs.

To consider charge-offs in the total default rate for each rating, we can use Prosper's reported
recovery rate of 9.5% ± 2.5%
[[2]](https://help.prosper.com/hc/en-us/articles/210013613-What-happens-if-a-borrower-misses-a-payment#:~:text=If%20a%20loan%20is%20more,range%20from%207%2D12%25.).

```{r}
# Adjust for charge-offs
loan_status_proportions_adjusted <- data.frame(
    COMPLETED_LO = loan_status_proportions[, "COMPLETED"] +
        0.07 * loan_status_proportions[, "CHARGEOFF"],
    COMPLETED_MID = loan_status_proportions[, "COMPLETED"] +
        0.095 * loan_status_proportions[, "CHARGEOFF"],
    COMPLETED_HI = loan_status_proportions[, "COMPLETED"] +
        0.12 * loan_status_proportions[, "CHARGEOFF"],
    DEFAULTED_LO = loan_status_proportions[, "DEFAULTED"] +
        0.93 * loan_status_proportions[, "CHARGEOFF"],
    DEFAULTED_MID = loan_status_proportions[, "DEFAULTED"] +
        0.905 * loan_status_proportions[, "CHARGEOFF"],
    DEFAULTED_HI = loan_status_proportions[, "DEFAULTED"] +
        0.88 * loan_status_proportions[, "CHARGEOFF"]
)
# Sort by `DEFAULTED_HI` ascending
loan_status_proportions_adjusted <- loan_status_proportions_adjusted[
                            order(loan_status_proportions_adjusted$DEFAULTED_HI), ]
round(loan_status_proportions_adjusted, digits = 4)
```

In the table above, `*_LO` represents the lower endpoint of recovery rate (7%), `*_MID` represents
the mid recovery rate (9.5%), and `*_HI` represents the upper endpoint of recovery rate (12.5%).

Adjusting for charge-offs, the data now show that `HR` loans outperform `E` loans, but all the
other ratings are in order.

For this sample, the default rates by rating are:

- AA: 2.9%-3.1%
- A: 7.7%-8.2%
- B: 15%-15.8%
- C: 23.4%-24.7%
- D: 31.1%-32.3%
- HR: 33.9%-35.8%
- D: 34%-35.9%

The defalt rates for `HR` and `D` loans are shockingly similar. Let's compare the average default
rates for the ratings to see which types are the best investments.

```{r}
# Group by Prosper rating
rating_groups <- split(df$borrower_rate, df$prosper_rating)

# Compute mean
average_rates <- tapply(df$borrower_rate, df$prosper_rating, mean)
(average_rates <- average_rates[, order(average_rates)]
```

TODO:

- Borrower rate vs term vs Amount borrowed
  - Hypothesis: if we have higher amount borrowed, we get longer term and lower rate
- Loan status vs borrower rate
  - Prosper should give better rates to completed loans
